---
title: "Group Project: CitiBike"
author: "Vinayak (\"Vinny Boi\") Ganesh and Jason Jin"
output: html_document
---

## Load the data

```{r, cache=TRUE}
citi <- read.csv("combined.csv")
weather <- read.csv("weather_data.csv")
```

## Clean the Citibike and weather date data

```{r, cache=TRUE}
library(dplyr)

weather <- mutate(weather, date_clean = as.Date(as.character(DATE), "%Y%m%d"))
citi <- mutate(citi, date_clean = as.Date(starttime, "%m/%d/%Y"))
citi <- mutate(citi, date_time = as.POSIXct(strptime(starttime, "%m/%d/%Y %H:%M:%S")))
```

## 3: Assymetric Traffic
Stations running out of bikes because of assymetric traffic (arrivals and departures are not equal, either arrival >> departure or departure >> arrival) is a big problem. Client would want to know which stations are candidates for increasing bike storage capacity. Client would like to see these stations on a map based visualization.

### Create bike inflow and outflow data

```{r, cache=TRUE}
library(plyr)

in_flow <- ddply(citi, c("end.station.id", "date_clean"), function(df) {
    nrow(df)
})
head(in_flow[in_flow$end.station.id == 72, ])

out_flow <- ddply(citi, c("start.station.id", "date_clean"), function(df) {
    nrow(df)
})
head(out_flow[out_flow$start.station.id == 72, ])

in_and_out_flow <- merge(in_flow, out_flow, by.x=c("end.station.id", "date_clean"), by.y=c("start.station.id", "date_clean"))
colnames(in_and_out_flow)[colnames(in_and_out_flow) == "end.station.id"] <- "station.id"
head(in_and_out_flow[in_and_out_flow$station.id == 72, ])
```

### Create average net flow data

```{r, cache=TRUE}
net_flow <- mutate(in_and_out_flow, net_flow = V1.x - V1.y)

head(net_flow[net_flow$station.id == 72, ])

net_flow_overall <- ddply(net_flow, c("station.id"), function(df) {
    mean(df$net_flow, na.rm = TRUE)
})
head(net_flow_overall)
```

### Merge average net flow with latitude/longitude data
```{r, cache=TRUE}
citi$start.station.latitude <- as.numeric(as.character(citi$start.station.latitude))
citi$start.station.longitude <- as.numeric(as.character(citi$start.station.longitude))
lat_long <- ddply(citi, c("start.station.id"), function(df) {
    c(df$start.station.latitude[1], df$start.station.longitude[1])
})
head(lat_long)

net_flow_overall_with_lat_long <- merge(net_flow_overall, lat_long, by.x=c("station.id"), by.y=c("start.station.id"))
colnames(net_flow_overall_with_lat_long) <- c("station_id", "avg_net_flow", "lat", "long")
net_flow_overall_with_lat_long$lat <- as.numeric(as.character(net_flow_overall_with_lat_long$lat))
net_flow_overall_with_lat_long$long <- as.numeric(as.character(net_flow_overall_with_lat_long$long))
head(net_flow_overall_with_lat_long)
```

### Get instances of negative and positive net flow separately

```{r, cache=TRUE}
neg_net_flow <- net_flow_overall_with_lat_long[net_flow_overall_with_lat_long$avg_net_flow < 0, ]
neg_net_flow <- mutate(neg_net_flow, magnitude = -1 * avg_net_flow)
neg_net_flow$lat <- as.numeric(as.character(neg_net_flow$lat))
neg_net_flow$long <- as.numeric(as.character(neg_net_flow$long))
head(neg_net_flow)

pos_net_flow <- net_flow_overall_with_lat_long[net_flow_overall_with_lat_long$avg_net_flow > 0, ]
pos_net_flow$lat <- as.numeric(as.character(pos_net_flow$lat))
pos_net_flow$long <- as.numeric(as.character(pos_net_flow$long))
head(pos_net_flow)
```

### Get latitude/longitude bounds
```{r, cache=TRUE}
lat <- c(min(neg_net_flow$lat[neg_net_flow$lat > 30]) - 0.01, max(neg_net_flow$lat[neg_net_flow$lat < 50]) + 0.01)
long <- c(min(neg_net_flow$long[neg_net_flow$long > -80]) -0.01, max(neg_net_flow$long[neg_net_flow$long < -60]) + 0.01)

lat
long
```

### Get map of NYC
```{r}
library(ggmap)

ggmap::register_google(key = "AIzaSyAIUHdzZd7CwFzO7WhtvUYi4l4ese9E4Lk")
ggmap::has_google_key()
nyc_base <- ggmap::get_map(location = c(mean(long), mean(lat)), zoom = 11)
```

### Plot negative net flow instances

```{r, cache=TRUE}
ggmap(nyc_base) +
    scale_x_continuous(limits = long, expand = c(0, 0)) +
    scale_y_continuous(limits = lat, expand = c(0, 0)) +
    geom_point(
        data = neg_net_flow,
        aes(x=long, y=lat, size=magnitude),
        color="red",
        alpha=0.5
    )
```

### Plot positive net flow instances

```{r, cache=TRUE}
ggmap(nyc_base) +
    scale_x_continuous(limits = long, expand = c(0, 0)) +
    scale_y_continuous(limits = lat, expand = c(0, 0)) +
    geom_point(
        data = pos_net_flow,
        aes(x=long, y=lat, size=avg_net_flow),
        color="darkgreen",
        alpha=0.5
    )
```

### Plot negative and positive net flow instances
```{r, cache=TRUE}
ggmap(nyc_base) +
    scale_x_continuous(limits = long, expand = c(0, 0)) +
    scale_y_continuous(limits = lat, expand = c(0, 0)) +
    geom_point(
        data = neg_net_flow,
        aes(x=long, y=lat, size=magnitude),
        color="red",
        alpha=0.5
    ) +
    geom_point(
        data = pos_net_flow,
        aes(x=long, y=lat, size=avg_net_flow),
        color="darkgreen",
        alpha=0.5
    )
```

## 4: Impact of Weather
The client wants to know the impact of weather (temperature, rain, snow, wind) on the CitiBike system.

### Merge Citibike and weather data

```{r, cache=TRUE}
merged <- merge(citi, weather, "date_clean")
```

### Impact of temperature, rain, snow, and wind on trip duration
```{r, cache=TRUE}
library(lmtest)

merged$tripduration <- as.integer(merged$tripduration)
tripduration_model <- lm(tripduration ~ PRCP + SNWD + SNOW + TMAX + TMIN + AWND, merged)
summary(tripduration_model)

library(ggplot2)

ggplot(merged, aes(x = TMAX, y = tripduration)) + geom_hist()
```

### Count number of trips per date by overall, user type, and gender
```{r}
library(plyr)

overall_counts <- ddply(merged, c("date_clean"), function(df) {
    nrow(df)
})
overall_counts <- merge(overall_counts, weather, "date_clean")

user_counts <- ddply(merged, c("date_clean", "usertype"), function(df) {
    nrow(df)
})
user_counts <- merge(user_counts, weather, "date_clean")

gender_counts <- ddply(merged, c("date_clean", "gender"), function(df) {
    nrow(df)
})
gender_counts <- merge(gender_counts, weather, "date_clean")
```

### Create generation column
```{r}
breaks <- c(0, 1964, 1979, 2000)
labels <- c("boomer", "gen_x", "millenials")

merged$birth.year <- as.numeric(merged$birth.year)
merged$gen <- cut(merged$birth.year, breaks, labels)
```

### Count number of trips per date by generation
```{r}
gen_counts <- ddply(merged[!is.na(merged$birth.year), ], c("date_clean", "gen"), function(df) {
    nrow(df)
})
gen_counts <- merge(gen_counts, weather, "date_clean")
```

### Impact of temperature, rain, snow, and wind on number of trips overall and number of trips by user type, gender, and generation

```{r}
overall_counts_model <- lm(V1 ~ PRCP + SNWD + SNOW + TMAX + TMIN + AWND, overall_counts)
summary(overall_counts_model)
```

```{r}
cust_counts_model <- lm(
    V1 ~ PRCP + SNWD + SNOW + TMAX + TMIN + AWND,
    user_counts[user_counts$usertype == "Customer", ]
)
summary(cust_counts_model)

sub_counts_model <- lm(
    V1 ~ PRCP + SNWD + SNOW + TMAX + TMIN + AWND,
    user_counts[user_counts$usertype == "Subscriber", ]
)
summary(sub_counts_model)
```

```{r}
gender_0_model <- lm(
    V1 ~ PRCP + SNWD + SNOW + TMAX + TMIN + AWND,
    gender_counts[gender_counts$gender == 0, ]
)
summary(gender_0_model)

gender_1_model <- lm(
    V1 ~ PRCP + SNWD + SNOW + TMAX + TMIN + AWND,
    gender_counts[gender_counts$gender == 1, ]
)
summary(gender_1_model)

gender_2_model <- lm(
    V1 ~ PRCP + SNWD + SNOW + TMAX + TMIN + AWND,
    gender_counts[gender_counts$gender == 2, ]
)
summary(gender_2_model)
```

```{r}
boomer_counts_model <- lm(
    V1 ~ PRCP + SNWD + SNOW + TMAX + TMIN + AWND,
    gen_counts[gen_counts$gen == "boomer", ]
)
summary(boomer_counts_model)

gen_x_counts_model <- lm(
    V1 ~ PRCP + SNWD + SNOW + TMAX + TMIN + AWND,
    gen_counts[gen_counts$gen == "gen_x", ]
)
summary(gen_x_counts_model)

millenials_counts_model <- lm(
    V1 ~ PRCP + SNWD + SNOW + TMAX + TMIN + AWND,
    gen_counts[gen_counts$gen == "millenials", ]
)
summary(millenials_counts_model)
```
